---
title: "pair-tutorial"
author: "Philip Berg"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{pair-tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

First we load the package:
```{r setup}
library(pair)
```
## Datasets included in `pair`
`pair` comes with two datasets, `ups` and `yeast` published in @berg2019evaluation (details can be found there).
The `ups` data is a global proteomics dataset without reducing peptides to proteins with *Chlamydomonas reinhardtii* (chlamy) peptides at the same concentration in all samples spiked-in with four different concentrations of the Universal Proteomics Standard Set 1 (UPS1).
The `yeast` data was generated by extracting reversibly oxidized cysteines in yeast or chlamy where the yeast extraction was added in two different concentrations and chlamy in the same in all samples.
The yeast data is therefore a lot noisier than the `ups`.
The yeast data looks as follows:
```{r, yeast}
yeast
```
where the `identifier` is a column that uniquely maps each cysteine to the observations in `ng50_*` or `ng100_*` (which are the relative concentrations the yeast cysteines were spiked in with) and the integer at the end indicates different repeated experiments.
The ups data looks as follows:
```{r, ups}
ups
```
where the `identifier` is as in the yeast dataset but uniquely maps peptides and `fmol25_*`, `fmol50_*`, and `fmol100_*` are different relative concentrations of UPS1 with different technical replicates.

## Data normalization
The first step is to normalize the data.
The function `plot_norm_box` generates boxplots of the raw data after log$_{2}$ transformation and two different normalization methods. If we first look at the yeast data, then the plot can be generated as follows:
```{r norm_vis_yeast, fig.width=7, fig.fullwidth=TRUE}
plot_norm_box(yeast, 'identifier')
```
The data is already has a stable median, but we still do normalization.
This can either be done `tmm` as introduced in @robinson2010scaling or with `psrn` as introduced in @anders2010differential.
To exemplify we use `psrn` and `tmm` as follows:
```{r, yeast_prnn}
yeast_psrn <- psrn(yeast, 'identifier')
yeast_tmm <- tmm(yeast)
```
Both `psrn` and `tmm` first normalizes the data and then log$_{2}$ transforms it unless the flag `log=FALSE` is used.
For the `ups` data we do the same thing:
```{r norm_vis_ups, fig.width=7, fig.fullwidth=TRUE}
plot_norm_box(ups, 'identifier')
```
we might notice that the x-axis labels are overlapping.
Since `plot_norm_box` returns a `ggplot` object, one can easily rotate the label by using the `+` syntax and `ggplot2::theme` as follows:
```{r norm_vis_ups_ang, fig.width=7, fig.fullwidth=TRUE}
plot_norm_box(ups, 'identifier') +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 15)
  )
```
in this way, any theme of the plot can easily be modified using `ggplot2` syntax.
One could also use the `plot_target` feature of `plot_norm_box` to select only a subset of samples to plot:
```{r norm_vis_ups_sub, fig.width=7, fig.fullwidth=TRUE}
plot_norm_box(ups, 'identifier', plot_target = contains('fmol25'))
```
which supports the `tidyselect` package operations for selecting columns.
Likewise, `tmm` and `psrn` have the flag `target` which can be used to select only a subset of columns to normalize.
If left unspecified they will use all numerical columns in the normalization!
They also contain the flag `load_info` which then makes the functions return a list with the normalized data, the estimated loading concentrations, and the estimated normalization factors used.
We then normalize the `ups` the same way as the `yeast`:
```{r, ups_prnn}
ups_psrn <- psrn(ups, 'identifier')
ups_tmm <- tmm(ups)
```

## Gamma regression for imputation and mean-variance normalization
From here on out, most functions in `pair` will depend on a design matrix.
A design matrix can be created as follows:
```{r, design}
yeast_design <- model.matrix(
  #Yeast has two conditions (1:2) and three replicates (each = 3)
  ~0+factor(rep(1:2, each = 3))
)
ups_design <- model.matrix(
  #ups has three conditions (1:3) and four replicates (each = 4)
  ~0+factor(rep(1:3, each = 4))
)
#The design matrices has to have the same name as the conditions:
colnames(yeast_design) <- paste0('ng', c(50, 100))
colnames(ups_design) <- paste0('fmol', c(25, 50, 100))
```
The gamma regression and mean-variance scatter plots can then be produced by `plot_gamma_regression`.
`plot_gamma_regression` plots the trends used for precision weights (later fed to `limma`'s `lmFit` similar to `voom`; @law2014voom) on the left and gamma regression used to infer the standard deviation given the mean (used for imputation) on the right.
For `yeast`:
```{r, gam_yeast, fig.width=7, fig.fullwidth=TRUE}
plot_gamma_regression(yeast_psrn, yeast_design, 'identifier')
```
and for `ups`:
```{r, gam_ups, fig.width=7, fig.fullwidth=TRUE}
plot_gamma_regression(ups_psrn, ups_design, 'identifier')
```
The two individual plots can be generated similarly by calling `plot_gamma_weights` or `plot_gamma_imputation`.
If one wants to inspect the estimated models that can be done similarly with `fit_gamma_regressions`, `fit_gamma_imputation`, or `fit_gamma_weights`.

## Data imputation and significance calling
### Multiple imputation
To perform multiple imputations and then run limma on each imputation can efficiently be performed by `run_pipeline`.
`run_pipeline` fits the needed gamma regressions and has a backend to `multidplyr` for parallel processing.
This can drastically increase the speed when running 1000 or more imputations.
Here is an example of the `yeast` data running 1000 imputations and using 5 parallel workers.
First, we need to generate a contrast matrix that indicates what comparisons to perform.
```{r, yeast_contras}
yeast_contrast <- limma::makeContrasts(
  contrasts = 'ng100-ng50',
  levels = yeast_design
)
```
This performs the test if the mean of the cysteines in the `ng100` conditions is different from the means in the `ng50` condition.
Further, it also means that the LFC will be calculated as `ng100 - ng50`.
We are now ready to run the imputation.
```{r, yeast_pipeline}
yeast_multi_imputation_results <- run_pipeline(
  data = yeast_psrn,
  design = yeast_design,
  contrast_matrix = yeast_contrast,
  imputations = 1000,
  workers = 5,
  id_col = 'identifier',
  plot_trend = FALSE # we have already looked at the mean-variance plots
)                    # otherwise, putting this flag to TRUE would generate those plots
```
`yeast_multi_imputation_results` will now be a `tibble` with each row containing one imputation and one decision from limma.
```{r, yeast_results}
yeast_multi_imputation_results
```
If one wants to access one particular imputation or decision that can be done with `$` and `[[` as follows.
```{r, yeast_example}
yeast_multi_imputation_results$imputed_data[[5]]
yeast_multi_imputation_results$limma_results[[5]]
```
which shows the results of the fifth imputation.
Next to perform p-value correction and binomial testing for the imputations we use `extract_results`.
```{r, yeast_extract}
yeast_reduced_results <- extract_results(
  data = yeast_psrn,
  results = yeast_multi_imputation_results,
  id_col = 'identifier',
  alpha = 0.05,
  abs_lfc = 1,
  pcor = 'fdr',
  null_hyp = 0.5
)
```

### Single imputation
To perform a single imputation one can use the `single_imputation` function.
```{r, yeast_sing_imp}
yeast_one_imputation <- single_imputation(yeast_psrn, yeast_design, 'identifier')
```
Next, we fit the gamma regression for estimating the precision weights with `fit_gamma_weights`.
```{r, yeast_gam}
yeast_gam_reg <- fit_gamma_weights(yeast_psrn, yeast_design, 'identifier')
```
We now run a single instance of `limma` and calculate the LFC with `run_limma_and_lfc`.
```{r, yeast_limma}
yeast_one_imputation_results <- run_limma_and_lfc(
  yeast_one_imputation,
  yeast_design, 
  yeast_contrast, 
  yeast_gam_reg, 
  'identifier'
)
```
One might want to perform some p-value correction on this data.
This can be one with `R`'s function `p.adjust`.
Further, since `run_limma_and_lfc` returns the data in tidy format and if there are several comparisons, `dplyr`'s `group_by` and `mutate` can performe one correction per conditions as follows:
```{r, yeast_adjust}
yeast_one_imputation_results <- yeast_one_imputation_results %>% 
  dplyr::group_by(comparison) %>% 
  dplyr::mutate(
    p_val = p.adjust(p_val, 'fdr')
  )
```
## Plotting the results
### Multiple imputation
The final decision for the analysis can be plotted with `plot_ma`.
`plot_ma` returns a `ggplot` object so it can easily be ascetically adjusted using `ggplot` syntax.
```{r, yeast_mult_ma, fig.width=7, fig.height=7, fig.fullwidth=TRUE}
plot_ma(yeast_reduced_results) + # Move legend to below the plot
  ggplot2::theme(
    legend.position = 'bottom'
  )
```


### Single imputation
For the single imputation, we need to include a few more inputs to `plot_ma`.
```{r, yeast_one_ma, fig.width=7, fig.height=7, fig.fullwidth=TRUE}
plot_ma(
  hits = yeast_one_imputation_results,
  data = yeast_one_imputation,
  id_col = 'identifier',
  alpha = .05,
  abs_lfc = 0.5
) + # Set color-blind friendly colors
  ggplot2::scale_color_manual(
    values = c('grey45', 'darkblue', 'darkred')
  )
```

# References
